A Fast and Low-Overhead Technique to Secure Programs Against Integer Overflows

### 2012-12-14 doc/Manuscript/CGO/paper.tex
Added final version of the paper to SVN

\documentclass{sigplanconf}[10pt]

% TODO: explain that we have not observed overflows in the safe instructions.
% TODO: explain that u-SSA does not yield noticeable performance improvements.

@@

\conferenceinfo{CGO '13}{February 23 - February 27. Shenzhen, China} 
\copyrightyear{2013}
\copyrightdata{[to be supplied]}

@@

\title{A fast and Low-Overhead Technique to Secure Programs Against
Integer Overflows}
%\subtitle{A case study of the Dinamica EGO toolkit}

@@

% TODO We need to explain better the wrapping-arithmetics, and we need to cite it
% properly. What could be a good (official) citation?

% Why there is still the need to devise efficient methods to eliminate them.
Programming languages such as Ada or Lisp can be customized to throw exceptions
whenever integer overflows are detected.
Furthermore, there exist recent works proposing to instrument binaries
derived from C, C++ and Java programs to detect the occurrence of overflows
dynamically~\cite{Brumley07,Dietz12}.
Thus, if an overflow happens, then the instrumented program can take some
action, such as to log the event, or to terminate the program.
However, this safety comes with a price: arithmetic operations need to be
surveilled, and the runtime checks cost time.
Zhang {\em et al.}~\cite{Zhang10} have proposed to eliminate some of this
overhead via a tainted flow analysis.
We have a similar goal, yet, our approach is substantially different.

% Our contribution: range analysis.
This paper describes the range analysis algorithm that
we have developed to eliminate overflow checks in instrumented programs.
As we show in Section~\ref{sec:range}, our algorithm has three core insights.
Firstly, in Section~\ref{sub:micro} we propose a three-phase approach to extract
information from comparisons between variables, e.g., $x < y$.
Previous algorithms either deal with these comparisons via
expensive relational analyses~\cite{Cousot78,Lakhdar11,Mine06}, or only
consider comparisons between variables and
constants~\cite{Mahlke01,Patterson95,Stephenson00}.
Secondly, in Section~\ref{sub:prop} we show how we rely on strongly connected
components to achieve speed and precision.
It is well-known that this technique is effective in speeding up constraint
resolution~\cite[Sec 6.3]{Nielson99}.
However, given our three-phase approach, we improve precision
by solving strong components in topological order.
Finally, in Section~\ref{sub:splitting} we propose a new program representation that
is more precise than other intermediate forms used to solve range
analysis sparsely.
This new live range splitting strategy is only valid if the instrumented
program terminates whenever an integer overflow is detected.
If we cannot rely on this guarantee, then our more conservative live range splitting
strategy produces the program representation that Bodik
{\em et al.}~\cite{Bodik00} call Extended Static Single Assignment
form.
In Section~\ref{sub:range} we show that an interprocedural implementation of our
algorithm analyzes programs with half-a-million lines of code in less than ten
seconds.
Furthermore, the speed and memory consumption of our range analysis grows linearly
with the program size.

% Our experimental results.
We use our range analysis to reduce the runtime overhead imposed by a dynamic
instrumentation library.
This instrumentation framework, a contribution of this paper that we describe in
Section~\ref{sec:dyn}, has been implemented in the LLVM compiler~\cite{Lattner04}.
We have logged overflows in a vast number of programs, and in this
paper we focus on SPEC CPU 2006.
We have re-discovered the integer overflows recently observed by
Dietz {\em et al.}~\cite{Dietz12}.
The performance of our instrumentation library, even without the support of range
analysis, is within the 5\% runtime overhead of Brumley {\em et al.}'s~\cite{Brumley07}
state-of-the-art algorithm.
The range analysis halves down this overhead.
Our static analysis algorithm avoids 24.93\% of the overflow
checks created by the dynamic instrumentation framework.
With this support, the instrumented SPEC programs are only 1.73\% slower.
Therefore, we show in this paper that securing programs against integer overflows
is very cheap.

\noindent
\textbf{The Danger that Integer Overflows Represent: }
Figure~\ref{fig:ex_buffer_overflow} illustrates an example of integer
overflow vulnerability.
This vulnerability allows an adversary to perform a buffer overrun attack
on a program that is apparently guarded against this type of exploit.
We will use the C datatype \texttt{char}, which represents
signed 8-bit numbers, so that the interested reader can easily
reconstruct this exploit.
Examples of actual vulnerabilities, with larger primitive types, are given
by Brumley {\em et al.}~\cite{Brumley07}, Dietz {\em et al.}~\cite{Dietz12},
and Zhang {\em et al.}~\cite{Zhang10}.
The function \texttt{read\_matrix} copies the linearized matrix \texttt{data}
into \texttt{buf}, to later process it, in line 12.
The arguments \texttt{w} and \texttt{h} determine the number of rows and
columns in the matrix.
If the \texttt{char} datatype could represent arbitrarily long numbers, then
this function would be safe against buffer overruns, due to the test at
line 3.
However, if, for instance, we have that \texttt{w}$= 6$ and \texttt{h}$= 22$,
then \texttt{buf\_size}$= -124$, due to the wrapping semantics of
integer arithmetics in C.
In this case, twelve words in the activation record of \texttt{read\_matrix}
would be overwritten.
By carefully crafting the contents of \texttt{data}, an adversary can, in this
way, change the return address of \texttt{read\_matrix}, diverting program's
execution to a system's function, such as \texttt{telnet}, for instance.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=\columnwidth]{images/ex_buffer_overflow}
\end{center}
\caption{\label{fig:ex_buffer_overflow}
An example of an exploitable integer overflow vulnerability.}
\end{figure}

@@

\subsection{The Interval Lattice}
\label{sub:lattice}

@@

We will use the program in Figure~\ref{fig:ex_eSSA_cgo}(a) to illustrate our
range analysis.
Figure~\ref{fig:ex_eSSA_cgo}(b) shows the same program in e-SSA
form~\cite{Bodik00},
and Figure~\ref{fig:ex_eSSA_cgo}(c) outlines the constraints that we extract
from this program.
There is a clear correspondence between instructions and constraints.
Our analysis is sparse~\cite{Choi91}; thus, we associate one, and only one,
constraint with each integer variable defined in the program.
A possible solution to the range analysis problem, as obtained via the
techniques that we will introduce in Section~\ref{sub:prop}, is given in
Figure~\ref{fig:ex_eSSA_cgo}(d).

@@

% The macro algorithm
Our range analysis algorithm works in a number of steps.
Firstly, we convert the program to an intermediate representation that gives
us subsidies to perform a sparse analysis.
We have tested our algorithm with two different representations, as we discuss
in Section~\ref{sub:splitting}.
Secondly, we extract constraints from the program representation.
Thirdly, we build a constraint graph, following the strategy pointed by
Su and Wagner~\cite{Su05}.
However, contrary to them, in a next phase we find the strongly connected
components in this graph, collapse them into super-nodes, and sort the
resulting digraph topologically.
Finally, for each strong component, we apply a three-phase approach to determine
the ranges of the variables.

@@

\noindent
\textbf{Widening: } we start solving constraints by determining how each
program variable might grow.
For instance, if a variable is only updated by sums with positive numbers, then
it only grows up.
If, instead, a variable is only updated by sums with negative numbers, then it
grows down.
Some variables can also grow in both directions.
We discover these growth patterns by abstractly interpreting the constraints
that constitute the strongly connected component.
We ensure termination via a widening operator.
Our implementation uses {\em jump-set widening}, which is typically used in
range analysis~\cite[p.228]{Nielson99}.
This operator is a generalization of Cousot and Cousot's original widening
operator~\cite{Cousot77}, which we describe below:
%
\begin{equation*}
I[Y] =
\begin{cases}
\mbox{if} \ I[Y] = [\bot, \bot] \ \mbox{then} \ e(Y) \\
\mbox{elif} \ \lb{e(Y)} < \lb{I[Y]}  \ \mbox{and} \ \ub{e(Y)} > \ub{I[Y]} \ \mbox{then} \ [-\infty, \infty] \\
\mbox{elif} \ \lb{e(Y)} < \lb{I[Y]} \ \mbox{then} \ [-\infty, \ub{I[Y]}] \\
\mbox{elif} \ \ub{e(Y)} > \ub{I[Y]} \ \mbox{then} \ [\lb{I[Y]}, \infty]
\end{cases}
\end{equation*}
%
We let $\lb{[l, u]} = l$ and $\ub{[l, u]} = u$.
We let $\bot$ denote non-itialized intervals, so that $[\bot, \bot] \sqcup
[l, u] = [l, u]$.
This operation only happens at $\phi$-nodes, because we evaluate constraints in
topological order.
The map $I$ and the abstract evaluation function $e$ are determined as in
Figure~\ref{fig:eval_function}.
We have an implementation of $e$ for each operation that the
target programming language provides.
Our current LLVM implementation has 18 different instances of $e$, including
signed and unsigned addition, subtraction, multiplication and division, plus
truncation, the bitwise integer operators and $\phi$-functions.

@@

\begin{figure}[t!]
\begin{center}
\begin{tabular}{ll}
Instruction & Dynamic Check \\ \\
$x = o_1 \ +_s \ o_2$ & $(o_1 > 0 \wedge o_2 > 0 \wedge x < 0) \ \ \vee$ \\
                      & $(o_1 < 0 \wedge o_2 < 0 \wedge x > 0)$ \\ \\
$x = o_1 \ +_u \ o_2$ & $x < o_1 \vee x < o_2$ \\ \\
$x = o_1 \ -_s \ o_2$ & $(o_1 < 0 \vee o_2 > 0 \vee x > 0) \ \ \vee$ \\
                      & $(o_1 > 0 \vee o_2 < 0 \vee x < 0)$ \\ \\
$x = o_1 \ -_u \ o_2$ & $o_1 < o_2$ \\ \\
$x = o_1 \ \times_{u/s} \ o_2$ & $x \neq 0 \Rightarrow x \div o_1 \neq o_2$ \\ \\
$x = o_1 \ \qside \ n$ & $(o_1 > 0 \wedge x < o_1) \vee (o_1 < 0 \wedge n \neq 0)$ \\ \\
$x = \ \downarrow_n \ o_1$ & $\mbox{cast}(x, \mbox{type}(o_1)) \neq o_1$ \\ \\
$x = \ (\mbox{type}_n) o_1$ & $\mbox{cast}(x, \mbox{type}(o_1)) \neq o_1$ \\
\end{tabular}
\end{center}
\caption{\label{fig:instrumentation}Overflow checks. We use $\downarrow_n$ for
the operation that truncates to $n$ bits, and we use $(\mbox{type}_n)$ for a
bit-cast to $n$ bits. Both operations have the same semantics in our setting.
The subscript $s$ indicates a signed instruction; the subscript $u$ indicate
an unsigned operation.}
\end{figure}

@@

\begin{figure}[t!]
\begin{center}
\begin{small}
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{lccccc}
& \textsc{Add} & \textsc{Sub} & \textsc{Mul} & \textsc{Shl} & \textsc{Trunc} \\
signed   & 12 & 12 & 6 & 8 & 3 \\
unsigned & 4  & 2  & 6 & 2 & 3
\end{tabular}
\end{small}
\caption{\label{fig:num_instructions}
Number of instructions used in each dynamic check.}
\end{center}
\end{figure}

@@

Which actions are performed once the overflow is detected depends on the
user, who has the option to overwrite the \texttt{handle\_ overflow}
function in Figure~\ref{fig:instrumented_cfg}.
Our library gives the user three options to handle overflows. 
First option: no-op.
This option allows us to verify the slowdown produced by the new instructions.
Second option: logging.
This is the standard option, and it preserves the behavior of the instrumented
program.
Whenever an overflow is detected, we print \texttt{Overflow detected in
FileName.cpp, line X.} in the standard error stream.
Third option: abort.
This option terminates the program once an overflow is detected.
Thus, it disallows undefined behavior due to integer overflows, and gives us the
opportunity to use the u-SSA form to get extra precision.

@@

The experiments also reveal that the memory consumption of our implementation
grows linearly with the program size.
Figure~\ref{fig:MemCorr} plots these two quantities.
The linear correlation, in this case, is even stronger than that found in
Figure~\ref{fig:TimeCorr}: the coefficient of determination is 0.9947.
The figure only shows our 100 largest benchmarks.
Again, SPEC \texttt{403.gcc} is the largest benchmark, requiring
265,588KB to run.
Memory includes stack, heap and the executable program code.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=\columnwidth]{images/MemCorr}
\end{center}
\caption{\label{fig:MemCorr}
Comparison between program size (number of var nodes in constraint
graphs) and memory consumption (KB).
Coefficient of determination = 0.9947.
}
\end{figure}

@@

%Prunning analysis
\begin{figure}[t!]
\begin{center}
\begin{small}
\begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}}|l|r|r|r|r|r|}
\hline
Benchmark  &   \#II &  \#E & \%(II, E) &  \#U & \%(II, U) \\ \hline
lbm        &  1,142 &      4 & 99.65\% &      4 & 99.65\% \\ \hline
milc       &  1,602 &  1,070 & 33.21\% &  1,065 & 33.52\% \\ \hline
namd       &  3,234 &  2,900 & 10.33\% &  2,900 & 10.33\% \\ \hline
dealII     & 36,157 & 29,870 & 17.39\% & 28,779 & 20.41\% \\ \hline
soplex     &  3,158 &  2,927 &  7.31\% &  2,897 &  8.26\% \\ \hline
h264ref    & 13,846 & 11,342 & 18.38\% & 11,301 & 18.08\% \\ \hline
astar      &    857 &    808 &  5.72\% &    806 &  5.95\% \\ \hline
sjeng      &  2,504 &  2,354 &  5.99\% &  2,190 & 12.54\% \\ \hline
mcf        &    165 &    164 &  0.61\% &    164 &  0.61\% \\ \hline
omnetpp    &  1,972 &  1,313 & 33.42\% &  1,313 & 33.42\% \\ \hline
gcc        & 18,669 & 15,282 & 18.14\% & 15,110 & 19.06\% \\ \hline
gobmk      & 14,129 & 12,563 & 11.08\% & 12,478 & 11.69\% \\ \hline
libquantum &    928 &    820 & 11.64\% &    817 & 11.96\% \\ \hline
bzip2      &  2,158 &  1,966 &  8.90\% &  1,966 &  8.90\% \\ \hline
hmmer      &  4,001 &  3,346 & 16.37\% &  3,304 & 17.42\% \\ \hline
Total      & 104,522 & 86,688 &        & 85,135 &         \\ \hline
\end{tabular*}
\end{small}
\end{center}
\caption{\label{fig:prunningTable}
Instrumentation library with support of static range analysis.
\#II: number of instructions that have been instrumented without range
analysis.
\#E: number of instructions instrumented in the e-SSA form program.
\#U: number of instructions instrumented in the u-SSA form program.}
\end{figure}

@@

\noindent
\textbf{Static Detection of Integer Overflows: }
Zhang {\em et al.}~\cite{Zhang10} have used static analysis to sanitize
programs against integer overflow based vulnerabilities.
They instrument integer operations in paths from a source to a sink.
In Zhang {\em et al.}'s context, sources are functions that read values from
users, and sinks are memory allocation operations.
Thus, contrary to our work, Zhang {\em et al.}'s only need to instrument
about 10\% of the integer operations in the program.
However, they do not use any form of range analysis to limit the number of
checks inserted in the transformed code.
Wang {\em et al.}~\cite{Wang09} have implemented a tool, IntScope, that combines symbolic execution and taint analysis to detect integer overflow vulnerabilities. The authors have been able to use this tool to successfully identify many vulnerabilities in industrial quality software. Our work, and Wang {\em et a.}'s work are essentially different: they use symbolic execution, whereas we rely on
range analysis.
Contrary to us, they do not transform the program to prevent or detect such event dynamically.
Still in the field of symbolic execution, Molnar
{\em et al.}~\cite{Molnar09} have implemented a tool, SmartFuzz, that analyzes
Linux x86 binaries to find integer overflow bugs.
They prove the existence of bugs by generating test cases for them.

\noindent
\textbf{Range Analysis and sparse representations:}
One of the goals of this paper is to describe a state-of-the-art range analysis
algorithm.
Many parts of this algorithm are well-known, others are novel.
The widening and narrowing operators from Section~\ref{sub:micro} were
introduced by Cousot and Cousot's seminal paper~\cite{Cousot77}.
The widening operator that we use in our implementation is well-known among
compiler writers~\cite[p.228]{Nielson99}.
The interval lattice from Section~\ref{sub:lattice} has been used in a
plethora of old and recent range analysis
solvers~\cite{Gough94,Gawlitza09,Logozzo10,Mahlke01,Oh12,Patterson95,Stephenson00,Su05}.
The use of futures to handle comparisons between variables is a contribution
of this paper.
It is well-known that the use of strongly connected components speeds up
constraint resolution systems~\cite[Sec 6.3]{Nielson99}.
However, publicly available implementations of range analysis such as those in
FLEX~\footnote{The MIT's FLEX/Harpoon compiler
provides an implementation of Stephenson's algorithm~\cite{Stephenson00}, and
is available at \texttt{http://flex.cscott. net/Harpoon/}.},
gcc~\footnote{Gcc's VRP pass (at \texttt{http://gcc.gnu.org/svn/gcc/trunk/gcc/
tree-vrp.c}) implements a variant of Patterson's
algorithm~\cite{Patterson95}.} or Mozilla's IonMonkey~\footnote{\texttt{hg.mozilla.org/projects/ionmonkey/file/629d1e6251b9/js/
src/ion/RangeAnalysis.cpp}. This code is implemented after Gough and
Klaeren~\cite{Gough94}} do not resort to this technique.
Furthermore, they only deal with comparisons between variables and constants.
The only work that we are aware of that has used range analysis to avoid
overflow checks is Sol {\em et al}'s~\cite{Sol11}; however, their algorithm
is much more limited than ours.
Contrary to us, Sol {\em et al.} eliminate overflow checks in
straight-line code traces produced by a just-in-time compiler, and they use the
runtime values of variables to improve their precision.

In this paper we provide a {\em sparse} implementation of range analysis.
Sparsity, in our context, means that we associate points in the lattice of
interest -- intervals in our case -- directly to variables.
Dense analyses map such information to pairs formed by variables and program
points.
Some previous analyses have achieved sparsity by using specific program
representations, like we did.
One of the program representations that we discuss in
Section~\ref{sub:splitting} -- the {\em
Extended Static Single Assignment} (e-SSA) form -- has been used by Bodik
{\em et al.}~\cite{Bodik00} and Gough {\em et al.}~\cite{Gough94}.
The other representation, which we call u-SSA, is very similar to
the \emph{Static Single Use} form (SSU).
There exist other variants of SSU~\cite{Lo98,Plevyak96,George03}.
The main difference to our representation is that our dataflow analysis
propagates information forwardly, whereas the SSU form has been conceived for
dataflow analyses that propagate information backwardly.
Thus, it requires special nodes at branches to merge information from
different program paths.
The SSU form programs are a subset of the programs in the so called
{\em Static Single Information} form, a program representation proposed by
Scott Ananian~\cite{Ananian99}.
SSI makes sparse the dataflow analyses that propagate information both
forwardly and backwardly.

@@

This paper has presented a static range analysis algorithm that reduces
the overhead necessary to secure programs against integer overflows.
This algorithm analyzes inter-procedurally programs with
half-a-million lines of code, i.e., almost one million constraints, in
ten seconds.
We proposed the notion of ``future bounds" to handle comparisons
between variables, and tested different program representations to improve our
precision.
Although the overhead of guarding programs against integer overflows is small,
as previous work has demonstrated, we believe that our technique is still
important, as some of these programs will be executed millions of times.
Our entire code plus the data that we show in this paper is publicly available.
A more extensive list of experiments is also available in our technical report.

\acks
We thank Douglas do Couto and Igor Costa for helping us with the LLVM code
base.
We also thank the anonymous reviewers for their useful comments and
suggestions.
This project is supported by FAPEMIG, grant 16899*01.

>>>>

\documentclass[10pt]{cgo13}

@@

\conferenceinfo{CGO '13}{23-27 February 2013, Shenzhen China.}
\copyrightyear{2013}
\copyrightdata{978-1-4673-5525-4/13/\$31.00 \copyright2013 IEEE}

@@

\title{A Fast and Low-Overhead Technique to Secure Programs Against
Integer Overflows}
%\subtitle{A case study of the Dinamica EGO toolkit}

@@

% Why there is still the need to devise efficient methods to eliminate them.
Programming languages such as Ada or Lisp can be customized to throw exceptions
whenever integer overflows are detected.
Furthermore, there exist recent works proposing to instrument binaries
derived from C, C++ and Java programs to detect the occurrence of overflows
dynamically~\cite{Brumley07,Dietz12}.
Thus, the instrumented program can take some action when an overflow happens,
such as to log the event, or to terminate the program.
However, this safety has a price: arithmetic operations need to be
surveilled, and the runtime checks cost time.
Zhang {\em et al.}~\cite{Zhang10} have eliminated some of this
overhead via a tainted flow analysis.
We have a similar goal, yet, our approach is substantially different.

% Our contribution: range analysis.
This paper describes the range analysis algorithm that
we have developed to eliminate overflow checks in instrumented programs.
As we show in Section~\ref{sec:range}, our algorithm has three core insights.
Firstly, in Section~\ref{sub:prop} we show how we rely on strongly connected
components to achieve speed and precision.
It is well-known that this technique is effective in speeding up constraint
resolution~\cite[Sec 6.3]{Nielson99}.
Yet, we go beyond: given our three-phase approach, we improve precision
by solving strong components in topological order.
Secondly, in Section~\ref{sub:micro} we describe this three-phase approach to
extract information from comparisons between variables, e.g., $x < y$.
Previous algorithms either deal with these comparisons via
expensive relational analyses~\cite{Cousot78,Lakhdar11,Mine06}, or only
consider comparisons between variables and
constants~\cite{Mahlke01,Patterson95,Stephenson00}.
Finally, in Section~\ref{sub:splitting} we propose a new program representation that
is more precise than other intermediate forms used to solve range
analysis sparsely.
This new live range splitting strategy is only valid if the instrumented
program terminates whenever an integer overflow is detected.
If we cannot rely on this guarantee, then our more conservative live range splitting
strategy produces the program representation that Bodik
{\em et al.}~\cite{Bodik00} call Extended Static Single Assignment
form.
In Section~\ref{sub:range} we show that an interprocedural implementation of our
algorithm analyzes programs with half-a-million lines of code in less than ten
seconds.
Furthermore, the speed and memory consumption of our range analysis grows linearly
with the program size.

% Our experimental results.
We use our range analysis to reduce the runtime overhead imposed by a dynamic
instrumentation library.
This instrumentation framework, which we describe in
Section~\ref{sec:dyn}, has been implemented in the LLVM compiler~\cite{Lattner04}.
We have logged overflows in a vast number of programs, and in this
paper we focus on SPEC CPU 2006.
We have re-discovered the integer overflows recently observed by
Dietz {\em et al.}~\cite{Dietz12}.
The performance of our instrumentation library, even without the support of range
analysis, is within the 5\% runtime overhead of Brumley {\em et al.}'s~\cite{Brumley07}
state-of-the-art algorithm.
The range analysis halves down this overhead.
Our static analysis algorithm avoids 24.93\% of the overflow
checks created by the dynamic instrumentation framework.
With this support, the instrumented SPEC programs are only 1.73\% slower.
Therefore, we show in this paper that securing programs against integer overflows
is very cheap.

@@

@@

We will use the program in Figure~\ref{fig:ex_eSSA_cgo}(a) to illustrate our
range analysis.
Figure~\ref{fig:ex_eSSA_cgo}(b) shows the same program in e-SSA
form~\cite{Bodik00},
and Figure~\ref{fig:ex_eSSA_cgo}(c) outlines the constraints that we extract
from this program.
There is a correspondence between instructions and constraints.
Our analysis is sparse~\cite{Choi91}; thus, it associates one, and only one,
constraint with each integer variable.
A possible solution to the range analysis problem, as obtained via the
techniques that we will introduce in Section~\ref{sub:prop}, is given in
Figure~\ref{fig:ex_eSSA_cgo}(d).

@@

% The macro algorithm
Our range analysis algorithm works in a number of steps.
Firstly, we convert the program to a representation that gives
us subsidies to perform a sparse analysis.
We have tested our algorithm with two different representations, as we discuss
in Section~\ref{sub:splitting}.
Secondly, we extract constraints from the program representation.
Thirdly, we build a constraint graph, following the strategy pointed by
Su and Wagner~\cite{Su05}.
However, contrary to them, in a next phase we find the strongly connected
components in this graph, collapse them into super-nodes, and sort the
resulting digraph topologically.
Finally, for each strong component, we apply a three-phase approach to determine
the ranges of the variables.

@@

\noindent
\textbf{Widening: } we start solving constraints by determining how each
program variable might grow.
For instance, if a variable is only updated by sums with positive numbers, then
it only grows up.
If, instead, a variable is only updated by sums with negative numbers, then it
grows down.
Some variables can also grow in both directions.
We discover these growth patterns by abstractly interpreting the constraints
that constitute the strongly connected component.
We ensure termination via a widening operator.
Our implementation uses {\em jump-set widening}, which is typically used in
range analysis~\cite[p.228]{Nielson99}.
This operator is a generalization of Cousot and Cousot's original widening
operator~\cite{Cousot77}, which we describe below:
%
\begin{equation*}
I[Y] =
\begin{cases}
\mbox{if} \ I[Y] = [\bot, \bot] \ \mbox{then} \ e(Y) \\
\mbox{elif} \ \lb{e(Y)} < \lb{I[Y]}  \ \mbox{and} \ \ub{e(Y)} > \ub{I[Y]} \ \mbox{then} \ [-\infty, \infty] \\
\mbox{elif} \ \lb{e(Y)} < \lb{I[Y]} \ \mbox{then} \ [-\infty, \ub{I[Y]}] \\
\mbox{elif} \ \ub{e(Y)} > \ub{I[Y]} \ \mbox{then} \ [\lb{I[Y]}, \infty]
\end{cases}
\end{equation*}
%
We let $\lb{[l, u]} = l$ and $\ub{[l, u]} = u$.
We let $\bot$ denote non-initialized intervals, so that $[\bot, \bot] \sqcup
[l, u] = [l, u]$.
This operation only happens at $\phi$-nodes, because we evaluate constraints in
topological order.
The map $I$ and the abstract evaluation function $e$ are determined as in
Figure~\ref{fig:eval_function}.
We have an implementation of $e$ for each operation that the
target programming language provides.
Our current LLVM implementation has 18 different instances of $e$, including
signed and unsigned addition, subtraction, multiplication and division, plus
truncation, the bitwise integer operators and $\phi$-functions.

@@

\begin{figure}[t!]
\begin{center}
\begin{tabular}{ll}
Instruction & Dynamic Check \\ \\
$x = o_1 \ +_s \ o_2$ & $(o_1 > 0 \wedge o_2 > 0 \wedge x < 0) \ \ \vee$ \\
                      & $(o_1 < 0 \wedge o_2 < 0 \wedge x > 0)$ \\ \\
$x = o_1 \ +_u \ o_2$ & $x < o_1 \vee x < o_2$ \\ \\
$x = o_1 \ -_s \ o_2$ & $(o_1 < 0 \vee o_2 > 0 \vee x > 0) \ \ \vee$ \\
                      & $(o_1 > 0 \vee o_2 < 0 \vee x < 0)$ \\ \\
$x = o_1 \ -_u \ o_2$ & $o_1 < o_2$ \\ \\
$x = o_1 \ \times_{u/s} \ o_2$ & $x \neq 0 \Rightarrow x \div o_1 \neq o_2$ \\ \\
$x = o_1 \ \qside \ n$ & $(o_1 > 0 \wedge x < o_1) \vee (o_1 < 0 \wedge n \neq 0)$ \\ \\
$x = \ \downarrow_n \ o_1$ & $\mbox{cast}(x, \mbox{type}(o_1)) \neq o_1$ \\
\end{tabular}
\end{center}
\caption{\label{fig:instrumentation}Overflow checks. We use $\downarrow_n$ for
the operation that truncates to $n$ bits.
The subscript $s$ indicates a signed instruction; the subscript $u$ indicate
an unsigned operation.}
\end{figure}

@@

\begin{figure}[t!]
\begin{center}
\begin{small}
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{lccccc}
& \textsc{Add} & \textsc{Sub} & \textsc{Mul} & \textsc{Shl} & \textsc{Trunc} \\
signed   & 12 & 12 & 6 & 8 & 3 \\
unsigned & 4  & 2  & 6 & 2 & 3
\end{tabular}
\end{small}
\caption{\label{fig:num_instructions}
Number of instructions used in each check.}
\end{center}
\end{figure}

@@

Which actions are performed once the overflow is detected depends on the
user, who has the option to overwrite the \texttt{handle\_overflow}
function in Figure~\ref{fig:instrumented_cfg}.
Our library gives the user three options to handle overflows. 
First option: no-op.
This option allows us to verify the slowdown produced by the new instructions.
Second option: logging.
This is the standard option, and it preserves the behavior of the instrumented
program.
Whenever an overflow is detected, we print \texttt{Overflow detected in
FileName.cpp, line X.} in the standard error stream.
Third option: abort.
This option terminates the program once an overflow is detected.
Thus, it disallows undefined behavior due to integer overflows, and gives us the
opportunity to use the u-SSA form to get extra precision.

@@

The experiments also reveal that the memory consumption of our implementation
grows linearly with the program size.
Figure~\ref{fig:MemCorr} plots these two quantities.
The linear correlation, in this case, is even stronger than that found in
Figure~\ref{fig:TimeCorr}: the coefficient of determination is 0.994.
The figure only shows our 100 largest benchmarks.
Again, SPEC \texttt{403.gcc} is the largest benchmark, requiring
265,588KB to run.
Memory includes stack, heap and the executable program code.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=\columnwidth]{images/MemCorr}
\end{center}
\caption{\label{fig:MemCorr}
Comparison between program size (number of var nodes in constraint
graphs) and memory consumption (KB).
Coefficient of determination = 0.994.
}
\end{figure}

@@

%Prunning analysis
\begin{figure}[t!]
\begin{center}
\begin{scriptsize}
\begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}}|l|r|r|r|r|r|}
\hline
Benchmark  &   \#II &  \#E & \%(II, E) &  \#U & \%(II, U) \\ \hline
lbm        &  1,142 &      4 & 99.65\% &      4 & 99.65\% \\ \hline
milc       &  1,602 &  1,070 & 33.21\% &  1,065 & 33.52\% \\ \hline
namd       &  3,234 &  2,900 & 10.33\% &  2,900 & 10.33\% \\ \hline
dealII     & 36,157 & 29,870 & 17.39\% & 28,779 & 20.41\% \\ \hline
soplex     &  3,158 &  2,927 &  7.31\% &  2,897 &  8.26\% \\ \hline
h264ref    & 13,846 & 11,342 & 18.38\% & 11,301 & 18.08\% \\ \hline
astar      &    857 &    808 &  5.72\% &    806 &  5.95\% \\ \hline
sjeng      &  2,504 &  2,354 &  5.99\% &  2,190 & 12.54\% \\ \hline
mcf        &    165 &    164 &  0.61\% &    164 &  0.61\% \\ \hline
omnetpp    &  1,972 &  1,313 & 33.42\% &  1,313 & 33.42\% \\ \hline
gcc        & 18,669 & 15,282 & 18.14\% & 15,110 & 19.06\% \\ \hline
gobmk      & 14,129 & 12,563 & 11.08\% & 12,478 & 11.69\% \\ \hline
libquantum &    928 &    820 & 11.64\% &    817 & 11.96\% \\ \hline
bzip2      &  2,158 &  1,966 &  8.90\% &  1,966 &  8.90\% \\ \hline
hmmer      &  4,001 &  3,346 & 16.37\% &  3,304 & 17.42\% \\ \hline
Total      & 104,522 & 86,688 &        & 85,135 &         \\ \hline
\end{tabular*}
\end{scriptsize}
\end{center}
\caption{\label{fig:prunningTable}
Instrumentation library with support of static range analysis.
\#II: number of instructions that have been instrumented without range
analysis.
\#E: number of instructions instrumented in the e-SSA form program.
\#U: number of instructions instrumented in the u-SSA form program.}
\end{figure}

@@

\noindent
\textbf{Static Detection of Integer Overflows: }
Zhang {\em et al.}~\cite{Zhang10} have used static analysis to sanitize
programs against integer overflow based vulnerabilities.
They instrument integer operations in paths from a source to a sink.
In Zhang {\em et al.}'s context, sources are functions that read values from
users, and sinks are memory allocation operations.
Thus, contrary to our work, Zhang {\em et al.}'s only need to instrument
about 10\% of the integer operations in the program.
However, they do not use any form of range analysis to limit the number of
checks inserted in the transformed code.
Wang {\em et al.}~\cite{Wang09} have implemented a tool, IntScope, that combines symbolic execution and taint analysis to detect integer overflow vulnerabilities. The authors have been able to use this tool to successfully identify many vulnerabilities in industrial quality software. Our work, and Wang {\em et al.}'s work are essentially different: they use symbolic execution, whereas we rely on
range analysis.
Contrary to us, they do not transform the program to prevent or detect such event dynamically.
Still in the field of symbolic execution, Molnar
{\em et al.}~\cite{Molnar09} have implemented a tool, SmartFuzz, that analyzes
Linux x86 binaries to find integer overflow bugs.
They prove the existence of bugs by generating test cases for them.

@@

This paper has presented a static range analysis algorithm that reduces
the overhead necessary to secure programs against integer overflows.
This algorithm analyzes inter-procedurally programs with
half-a-million lines of code, i.e., almost one million constraints, in
ten seconds.
We proposed the notion of ``future bounds" to handle comparisons
between variables, and tested different program representations to improve our
precision.
Although the overhead of guarding programs against integer overflows is small,
as previous work has demonstrated, we believe that our technique is still
important, as some of these programs will be executed millions of times. \\
\noindent
\textbf{Software: } Our implementation is
publicly available at \\
\texttt{http://code.google.com/p/range-analysis/}.

\acks
We thank Douglas do Couto and Igor Costa for helping us with the LLVM code
base.
We also thank the anonymous reviewers for comments and
suggestions.
This project is supported by FAPEMIG and Intel.
